Master Thesis

### 2013-05-13 ig.tex
Talked about Frigram a little

The way the polarities are handled is that every model (output syntactic
tree) is required to have only saturated polarities on its features
(i.e. no positive, negative or virtual polarities). Whenever more than
one node of the PTD is mapped onto a single node of the syntactic tree,
the polarities of each feature are combined. The combination mechanism
allows to combine a positively polarized feature with a negatively
polarized one to yield a saturated one. Virtual features can be factored
out only by combining with a saturated polarity.

If we look at the PTDs of Figure~\ref{fig:ig-eptds}, we can see this
polarity mechanism in action. The EPTD of $Jean$ has a positive $cat$
feature in its root node saying that it provides one $np$, and a
negative $funct$ feature saying that it expects some function. In the
EPTD of $voit$, we have two nodes for the subject and object, both of
them expecting an $np$ and providing them the $subj$ and $obj$
functions, respectively. The root of the EPTD than provides a complete
sentence of category $s$ and expects some function that this sentence
will play. The full stop EPTD finalizes the sentence by accepting a node
with category $s$ and giving it a $void$ function.

@@

Now that we understand the constraints imposed by PTDs, we can start to
see that the syntactic tree in Figure~\ref{fig:ig-parse} is truly a
model of the PTDS given in Figure~\ref{fig:ig-eptds} and more than that,
it is the only minimal model. To illustrate more clearly how the
polarities and constraints of the individual PTDs end up generating the
syntactic tree of Figure~\ref{fig:ig-parse}, we can look at the result
of superposing the EPTDs of $la$ and $voit$ by merging the nodes (1,2)
and (1,5) with (2,2) and (2,4) respectively on
Figure~\ref{fig:ig-partial}.

@@

>>>>

The way the polarities are handled is that every model (output syntactic
tree) is required to have only saturated polarities on its features
(i.e. no positive, negative or virtual polarities). Whenever more than
one node of the PTD is mapped onto a single node of the syntactic tree,
the polarities of each feature are combined. The combination mechanism
allows to combine a positively polarized feature with a negatively
polarized one to yield a saturated one. Virtual features can be factored
out only by combining with a saturated polarity (they are somewhat
analogous to the $=_{c}$ constraints of LFG \cite{kaplan1982lexical}).

If we look at the PTDs of Figure~\ref{fig:ig-eptds}, we can see this
polarity mechanism in action. The EPTD of $Jean$ has a positive $cat$
feature in its root node saying that it provides one $np$, and a
negative $funct$ feature saying that it expects some function. In the
EPTD of $voit$, we have two nodes for the subject and object, both of
them expecting an $np$ and providing them the $subj$ and $obj$
functions, respectively. The root of the EPTD then provides a complete
sentence of category $s$ and expects some function that this sentence
will play. The full stop EPTD finalizes the sentence by accepting a node
with category $s$ and giving it a $void$ function.

@@

Now that we understand the constraints imposed by PTDs, we can start to
see that the syntactic tree in Figure~\ref{fig:ig-parse} is truly a
model of the PTDs given in Figure~\ref{fig:ig-eptds} and more than that,
it is the only minimal model. To illustrate more clearly how the
polarities and constraints of the individual PTDs end up generating the
syntactic tree of Figure~\ref{fig:ig-parse}, we can look at the result
of superposing the EPTDs of $la$ and $voit$ by merging the nodes (1,2)
and (1,5) with (2,2) and (2,4) respectively on
Figure~\ref{fig:ig-partial}.

@@

\subsubsection{Frigram, an interaction grammar}

The reason we are interested in interaction grammars is because of the
existence of Frigram, a large scale grammar of French which is
lexicalized by Frilex, an independent lexical resource that we will be
building our grammar on as well. Furthermore, as we will see
in~\ref{sssec:link-ig-acg}, interaction grammars are closely linked with
abstract categorial grammars, which makes Frigram a suitable grammar to
use as a guide when developing our grammar.

In this section, we will introduce the metagrammatical structure of
Frigram and talk about how it uses the formalism of interaction grammars
to solve several tricky linguistic phenomena.

Frigram is a wide-coverage interaction grammar of French. As we
mentioned before, an interaction grammar is a set of elementary
polarized tree descriptions. In a lexicalized interaction grammar, each
of these EPTDs is connected to a specific wordform whose particular use
it describes. Given the amount of wordforms that a useable grammar of
French would need to cover, the number of EPTDs in our grammar could
easily reach hundreds of thousands, if not millions.

The first step in fighting this explosion is to factor out irrelevant
differences between similar wordforms. For this purpose, Frilex was
created. Frilex is a morpho-syntactic lexicon of French compiled from
various other pre-existing lexicons of French. It is in effect a large
relation linking the wordforms of French to \emph{hypertags}, feature
structures describing the morphological properties of the wordforms and
their syntactic valencies.

With Frilex in place, Frigram can be defined as a set of unanchored
EPTDs which are paired with feature structures which delimit the subset
of the Frilex items to which the EPTDs apply. This kind of simple
``metagrammar'' already saves us a lot of effort, the number of
unanchored EPTDs defined by Frigram is somewhere around 4000 (TODO: Cite
the Frig documentation somehow).

However, defining some 4000 EPTDs manually still seems like a very
tedious, error prone and hard to maintain approach. Many of the EPTDs
have to repeatedly describe common phenomena such as subject-object
agreement or predicate-argument saturation. Similarly as in other
software endeavors, it would be preferable to define these common
patterns in some reusable module and compose EPTDs from these building
blocks. This is where XMG steps in.

XMG is a metagrammar compiler adding yet another level of indirection
between what the grammar author writes and what ends up in the
bottom-level interaction grammar. XMG provides a language that lets the
grammar author define the lexical items of his grammar and to combine
these definitions to yield new and more elaborate lexical items.

In the case of Frigram, the lexical items (termed \emph{classes} in XMG)
take the shape of PTDs coupled with feature structures which
circumscribe the interface to Frilex. Both tree descriptions and feature
structures have very natural ways of composing together, by conjunction
and unification, respectively. This, alongside with disjunction, are the
chief tools that the grammar authors can use (and in the case of
Frigram, have used) to succintly define their grammar.

With the added capability of composing classes, the definition of
Frigram reduces to about 400 class definitions. Of these classes, 160
are terminal, meaning that they actually describe lexical items that are
to be included in the final grammar. Thanks to the disjunction
composition operator of XMG, these 160 classes end up generating the
staggering 4000 EPTDs.

\subsubsection{The link with abstract categorial grammars}
\label{sssec:link-ig-acg}
